Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-01 content/contribution/continuous-datalog.tex
Fixed minor bugs

	In regards to Continuous Datalog, an ``instant'', a ``stream of facts'' and a ``sliding window'' over a stream are defined as follows:

\begin{definition}[Instant]\label{def:continuous datalog: instant}
An instant is a discrete point in time, such that an instant ${i \in \mathbb{N}}$.

@@

\begin{definition}[Fact Instance]\label{def:continuous datalog: fact instance}
A fact instance is a combination of the signature of a fact (its
predicate $p$, subject $s$ and object $o$) and its entailment time
$t$; a fact instance is denoted ${p(s,o):t}$, and is unique.
\end{definition}

\begin{definition}[Stream of Facts]\label{def:continuous datalog: stream}
A stream of facts is a \emph{sequence} of Datalog facts ordered by
some temporal characteristic of each instance of each fact, e.g. the
time at which each was entailment by some process at the source of the
stream.  The distinction between instances of facts within a
stream is represented in Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

\begin{definition}[Sliding Window]\label{def:continuous datalog: window}
A sliding window of range ${X \in \mathbb{N}}$ over any stream $S$ at
any given instant ${i \in \mathbb{N}}$ is the \emph{set} of instances
of facts that occur in $S$ during the interval ${(i-X,i]}$, and is
  denoted ${W^{S,X}_{i}}$.
\end{definition}

@@

\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over any stream $S$ at any instant 
${i \in \mathbb{N}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of any range ${X \ge 0}$ over any stream $S$ at any
given instant ${i \in \mathbb{N}}$ is equal to the union of the
$X$ consecutive windows of range 1 over the stream $S$ preceding and
including that at instant $i$.
\begin{equation*}
W^{S,X}_{i} = \mathop{\cup}_{i=0}^{X-1} W^{S,1}_{i-i}
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of any range ${X \in \mathbb{N}}$ over any stream $S$ at any
instant ${i \in \mathbb{N}}$ is disjoint from a second window of any
range ${Y \in \mathbb{N}}$ over $S$ at any instant after ${i - X}$.
\begin{equation*}
\forall X' \geq X \, \left( W^{S,X}_{i} \cap W^{S,Y}_{i-X'} = \varnothing \right)
\end{equation*}
\end{axiom}

	Traditionally, a set of axioms $A$ entails another set of axioms $E$ in which the existence of one fact in $E$ may be justified by many distinct subsets of $A$.
	When considering streams of facts, in which one fact may occur multiple times and be considered distinct based on its entailment time, temporal information would be lost if sets of axioms including fact instances (denoted $A_{inst}$) were to entail sets of axioms without.
	As such, where a fact would be entailed by some set of facts $A$ that are instantiated on some stream, an instance of that fact should be entailed by the set of instances of those facts present on that stream.
	Furthermore, each such justification in $A_{inst}$ for a fact entails a unique instance of that fact, such that each entailed instance has a single justification within the original set of axioms, and many instances of the entailed fact may occur in the entailed set of axioms $E_{inst}$.
	This form of entailment shall be henceforth referred to as ``instance entailment'', and shall be the norm when discussing the entailments of sets of axioms that include instances of facts.

	It should be noted that removing all temporal information from a set of axioms including fact instances $A_{inst}$ produces the set of traditional Datalog axioms $A$ instantiated in $A_{inst}$.
	This is be represented as the function ${trad\left( A_{inst} \right) = A}$.
	In addition, applying the ${trad()}$ function to both the original axiom instances $A_{inst}$ and the entailed axiom instances $E_{inst}$ maintains the relationship of entailment between the two sets:
	\[ \left( A_{inst} \vDash E_{inst} \right) \rightarrow \left( A \vDash E \right) \]

	From this description of \emph{instance entailment}, Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be derived.

	\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
		If two sets, $A_{inst}$ and $B_{inst}$, of axiom instances are disjoint, then the any set $E_{inst}$ of axiom instances each justified at least in part by $A_{inst}$ is necessarily disjoint from any set $F_{inst}$ of axiom instances each justified at least in part by $B_{inst}$, because entailed fact instances are distinct with regards to the fact instances that justify them.
		\begin{multline*}
			\left( A_{inst} \cap B_{inst} = \varnothing \right) \wedge \left( A_{inst} \vDash E_{inst} \right) \\
				\wedge \left( B_{inst} \vDash F_{inst} \right) \rightarrow \left( E_{inst} \cap F_{inst} = \varnothing \right)
		\end{multline*}
	\end{axiom}

@@

	\begin{definition}[Continuous Datalog Program]\label{def:continuous datalog: CDP}
		A Continuous Datalog Program (${CDP}$) is composed of a set of Positive Datalog axioms $A$, a set of streams of facts $S$ and the range ${R \in \mathbb{N}}$ of the windows over each stream in $S$.
		This specifies that the streamed data from $S$ is to be considered valid from its arrival for the duration represented by $R$.
		Such a continuous program entails a triple composed of the traditional entailments of ${A \vDash A'}$, a stream of entailed instances $P$ and a stream of instance negations $N$, whose contents are defined in Axioms~\ref{axiom:continuous datalog: positive window increment},~\ref{axiom:continuous datalog: negative window increment} and~\ref{axiom:continuous datalog: entailment precedes negation}.
		Streams $P$ and $N$ are therefore the continuous reasoning equivalents of the streams produced by the \emph{IStream} and \emph{DStream} operators of the \emph{CQL} continuous query language \citep{CQL} respectively.
		The truth value of a ${CDP}$ is instantaneous, with the the truth function of Continuous Datalog $T^{CD}$ taking both a program ${CDP}$ and an instant ${i \in \mathbb{N}}$ as arguments.
		\begin{equation*}
			CDP = \{A,S,R\} \vDash \{A',P,N\}
		\end{equation*}
	\end{definition}

	\begin{definition}[State of a ${CDP}$]\label{def:continuous datalog: CDPt}
		The state of a ${CDP}$ at a given instant ${i \in \mathbb{N}}$ (${CDP_{i}}$) is the union of the axioms $A$ and the window $W^{S,R}_{i}$.
		As both $A$ and ${trad\left( W^{S,R}_{i} \right)}$ are sets of Datalog axioms, a ${CDP_{i}}$ constitutes a valid Datalog program that may be evaluated by the truth function of Datalog $T^{D}$, the result of which is the definition for the instantaneous truth of the ${CDP}$ at instant $i$ by $T^{CD}$.
		The instantaneous entailments of a $CDP_{i}$ are the union of the persistent entailments of ${A \vDash A'}$ (which are guaranteed to be entailed by every state, as Positive Datalog is monotonic) and the set $E_{i}$ of instances of the transient entailments that are justified at least in part by some subset of $W^{S,R}_{i}$.
		\begin{equation*}
			T^{CD} \left( CDP, i \right) = T^{D} \left( CDP_{i} \right)
		\end{equation*}
		\begin{equation*}
			CDP_{i} = A \cup W^{S,R}_{i} \vDash A' \cup E_{i}
		\end{equation*}
	\end{definition}

@@

	\begin{axiom}\label{axiom:continuous datalog: positive window increment}
		The window of range 1 at any given instant ${i \in \mathbb{N}}$ over the stream $P$ entailed by a ${CDP}$ is defined as the set of facts that are entailed by the state of the ${CDP}$ at instant $i$ that were not entailed by the state of ${CDP}$ at the previous instant instance ${i-1}$.
		\begin{equation*}
			W^{P,1}_{i} = \left( A' \cup E_{i} \right) \setminus \left( A' \cup E_{i-1} \right) = E_{i} \setminus E_{i-1}
		\end{equation*}
	\end{axiom}

	\begin{axiom}\label{axiom:continuous datalog: negative window increment}
		The window of range 1 at any given instant ${i \in \mathbb{N}}$ over the stream $N$ entailed by a ${CDP}$ is defined as the set of facts that were entailed by the state of the ${CDP}$ at the previous instant instance ${i-1}$ that are no longer entailed by the state of ${CDP}$ at instant $i$.
		\begin{equation*}
			W^{N,1}_{i} = \left( A' \cup E_{i-1} \right) \setminus \left( A' \cup E_{i} \right) = E_{i-1} \setminus E_{i}
		\end{equation*}
	\end{axiom}

	\begin{axiom}\label{axiom:continuous datalog: entailment precedes negation}
		Facts may only be negated by some part of their justification leaving the sliding window $W^{S,R}_{i}$, this necessarily being after they were entailed.
		As such, a window of any range ${X \in \mathbb{N}}$ over the stream of entailments $P$ at any given instant ${i \in \mathbb{N}}$ is disjoint from a window of any range ${Y \in \mathbb{N}}$ over the the stream of entailment negations $N$ at any instant after ${i - X}$
		\begin{equation*}
			\forall X' \geq X \left( W^{P,X}_{i} \cap W^{N,Y}_{i-X'} = \varnothing \right)
		\end{equation*}
	\end{axiom}

	From this, we can see that each instance of an entailment in some $E_{i}$ may take the same form as an instance of any other fact in a stream of facts, being ${p(s,o):t}$, where $t$ is the real time associated with the instant $i_{e}$ in which ${p(s,o)}$ becomes newly justified by the ${CDP}$
	(more specifically, by some subset of the instances in $W^{S,R}_{i_{e}}$).
	As such, the stream of entailment instances $P$ of a ${CDP}$ takes the same form as the \emph{streams of facts} already described, thus producing a concise stream of entailments that may be read by further stream processing.
	In addition, the stream of negations $N$ of a ${CDP}$ also takes this form, but differs from the streams discussed thus far in that it is not ordered by the entailment time $i_{e}$ of the entailment instances but by the time at which their justification \emph{ceases} to be valid
	(i.e. the first instant ${i_{n} \geq i_{e}}$ in which $W^{S,R}_{i_{n}}$ no longer contains the subset of fact instances first present in $W^{S,R}_{i_{e}}$ that justified the entailment instance ${p(s,o):t}$).

	Thus far the translation from instantaneous sets of entailments to streams of entailments has been stated axiomatically as the definition of those streams.
	However, it can be shown from these axioms,
	as in \ref{sec:semantics: proof for entailments},
	that it is also possible to reconstruct exactly the instantaneous set $E_{i}$ of entailment instances of any ${CDP}$ at any given instant from the windows of sufficient range over its entailed streams $P$ and $N$ at that instant.
	\[ \forall Y \geq X \geq R \, \left( E_{i} = W^{P,X}_{i} \setminus W^{N,Y}_{i} \right) \]
	The minimum range $X$ is such to achieve the \emph{complete} set of entailment instances $E_{i}$ of the ${CDP}_{i}$.
	In addition, the minimum range $Y$ is such to achieve the \emph{correct} set $E_{i}$.
	As such, a \emph{valid window} that contains only and exactly those instances of entailments in $E_{i}$ at every instant $i$ may be trivially maintained from the streams $P$ and $N$ continuously entailed by some ${CDP}$.
\end{nestedsection}

>>>>

%% fixme
%	In regards to Continuous Datalog, an ``instant'', a ``stream of facts'' and a ``sliding window'' over a stream are defined as follows:

\begin{definition}[Datalog Program]
\label{def:continuous datalog: datalog program}
A conventional Datalog program $\Pi = \langle R, F\rangle$ is a tuple
which comprises a finite set of rules $R$ and a finite set of facts
$F$. Facts are extensional atomic formulae of first-order logic, and
rules are of the form:

\[\alpha_0 :- \alpha_1 , \ldots , \alpha_n\]

\noindent where $\alpha_0$ is an intensional atomic formula and
$\alpha_1 \ldots \alpha_n$ are a mixture of intensional and
extensional atomic formulae.

For the purpose of defining Continuous Datalog in the context of the
Semantic Web, we treat facts as 3-tuples $\langle s, p, o \rangle$,
where $s, p, o$ are respectively the subject, predicate and object of
an RDF triple. 
\end{definition}

\begin{definition}[Instant]
\label{def:continuous datalog: instant}
An instant is a discrete point in time, such that an instant 
${i \in \mathbb{N}}$.

@@

\begin{definition}[Fact Instance]
\label{def:continuous datalog: fact instance}
A {\em fact instance} is a fact with some additional temporal
annotation; it is the combination of a fact's signature (its predicate
$p$, subject $s$ and object $o$) and a timestamp $t$ (an instant) that
is used to record the time at which the fact was asserted or
entailed. A fact instance is denoted $\langle s, p, o \rangle \! : \!
t$, and is unique.

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function $strip$ which maps a fact instance to its underlying fact.
\end{definition}

\begin{definition}[Stream of Facts]
\label{def:continuous datalog: stream}
A stream of facts $S$ is a \emph{sequence} of fact instances ordered
by their timestamps; the distinction between multiple instances of a
fact within a stream is represented in 
Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

\begin{definition}[Sliding Window]
\label{def:continuous datalog: window}
A sliding window of range ${x \in \mathbb{N}}$ over any stream $S$ at
any given instant ${i \in \mathbb{N}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, and is
  denoted ${W^{S,x}_{i}}$.
\end{definition}

@@

\begin{axiom}\label{axiom:continuous datalog: window range leq 0}
A window of range $0$ over a stream $S$ at an instant 
${i \in \mathbb{N}}$ is always empty.
\begin{equation*}
W^{S,0}_{i} = \varnothing
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window composition}
A window of range ${x \ge 0}$ over a stream $S$ at an instant 
${i \in \mathbb{N}}$ is equal to the union of the $x$ consecutive
windows of range $1$ over the stream $S$ preceding and including that at
instant $i$.
\begin{equation*}
W^{S,x}_{i} = \mathop{\cup}_{j=0}^{x-1} W^{S,1}_{i-j}
\end{equation*}
\end{axiom}

\begin{axiom}\label{axiom:continuous datalog: window disjointness}
A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an
instant ${i \in \mathbb{N}}$ is disjoint from a second window of 
range ${y \in \mathbb{N}}$ over $S$ at any instant before $i - x$.

\begin{equation*}
W^{S,x}_{i} \cap W^{S,y}_{i-z} = \varnothing
\end{equation*}

\noindent where $z \geq x$.
\end{axiom}

In Definition~\ref{def:continuous datalog: datalog program}, we took a
Datalog program to consist of a set of rules and a set of facts.
Taken together, the set of facts $F$ entails another set of factss $E$
in which the existence of one fact in $E$ may be justified by many
distinct subsets of $F$.

When considering streams of facts, in which there may be many distinct
instances of a given fact, temporal information would be lost if sets
of fact instances (denoted $F_{inst}$) were to entail sets of plain
facts.  As such, where a fact $f$ would be entailed by some set of
facts $F$ that are instantiated on some stream, an instance of $f$
should be entailed by the set of instances of those facts present on
that stream.  Furthermore, each such justification in $F_{inst}$ for a
fact entails a unique instance of that fact, such that each entailed
instance has a single justification within the original set of facts,
and many instances of the entailed fact may occur in the entailed set
of facts $E_{inst}$.  This form of entailment shall be henceforth
referred to as {\em instance entailment}, and shall be the norm when
discussing the entailments of sets of fact instances.

It should be noted that removing all temporal information from a set
of fact instances $F_{inst}$ produces the set of facts $F$
instantiated in $F_{inst}$:

\[F = \{ strip(x) | x \in F_{inst} \}\]

In addition, applying the ${strip()}$ function to both the original
fact instances $F_{inst}$ and the entailed fact instances $E_{inst}$
maintains the relationship of entailment between the two sets:

\[ \left( F_{inst} \vDash E_{inst} \right) \rightarrow \left( F \vDash E \right) \]

From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:

\begin{axiom}
\label{axiom:continuous datalog: entailment disjointness}

If two sets of fact instances, $A_{inst}$ and $B_{inst}$, are
disjoint, then a set $F_{inst1}$ of entailed fact instances (each of
which is justified at least in part by $A_{inst}$) is necessarily
disjoint from any set $F_{inst2}$ of entailed fact instances (each
justified at least in part by $B_{inst}$), because entailed fact
instances are distinct with regards to the fact instances that justify
them.

\begin{multline*}
\left( A_{inst} \cap B_{inst} = \varnothing \right) \wedge \left( A_{inst} \vDash E_{inst} \right) \\
\wedge \left( B_{inst} \vDash F_{inst} \right) \rightarrow \left( E_{inst} \cap F_{inst} = \varnothing \right)
\end{multline*}
\end{axiom}

@@

\begin{definition}[Continuous Datalog Program]
\label{def:continuous datalog: CDP}

A Continuous Datalog Program (CDP) is a tuple $\langle R, F, S, r
\rangle$ which comprises a finite set of rules $R$, a finite set of
facts $F$, a stream of facts $S$ and a range $r \in \mathbb{N}$ for
the window over $S$. This specifies that the streamed data from $S$ is
to be considered valid from its arrival for the duration represented
by $r$.

The entailment of such a program is composed of a stream of entailed
instances $P$ and a stream of instance negations $N$, whose contents
are defined in 
Axioms~\ref{axiom:continuous datalog: positive window increment}, 
\ref{axiom:continuous datalog: negative window increment} and
\ref{axiom:continuous datalog: entailment precedes negation}. 
The streams $P$ and $N$ are therefore the continuous
reasoning equivalents of the streams produced by the \emph{IStream}
and \emph{DStream} operators of the \emph{CQL} continuous query
language \citep{CQL} respectively.  The truth value of a CDP is
instantaneous, with the the truth function of Continuous Datalog
$T^{CD}$ taking both a program ${CDP}$ and an instant ${i \in
  \mathbb{N}}$ as arguments.
\end{definition}

\begin{definition}[State of a CDP]
\label{def:continuous datalog: CDPt}

The state of a CDP at a given instant $i$ (denoted as ${CDP_{i}}$) is
the union of the set of facts $F$ and the window $W^{S,r}_{i}$.  As both $F$
and ${strip( W^{S,r}_{i} )}$ are sets of Datalog axioms, 
${CDP_{i}}$ constitutes a valid Datalog program that may be evaluated
by the truth function of Datalog $T^{D}$, the result of which is the
definition for the instantaneous truth of the ${CDP}$ at instant $i$
by $T^{CD}$.  The instantaneous entailments of a $CDP_{i}$ are the
union of the set $E$, the persistent entailments of $F$ (which are
guaranteed to be entailed by every state, as Positive Datalog is
monotonic), and the set $E_{i}$ of instances of the transient
entailments that are justified at least in part by some subset of
$W^{S,r}_{i}$.

\begin{equation*}
T^{CD} \left( CDP, i \right) = T^{D} \left( CDP_{i} \right)
\end{equation*}
\begin{equation*}
CDP_{i} = F \cup W^{S,r}_{i} \vDash E \cup E_{i}
\end{equation*}
\end{definition}

@@

\begin{axiom}
\label{axiom:continuous datalog: positive window increment}

The window of range 1 at any given instant ${i \in \mathbb{N}}$ over
the stream $P$ entailed by a ${CDP}$ is defined as the set of facts
that are entailed by the state of the ${CDP}$ at instant $i$ that were
not entailed by the state of ${CDP}$ at the previous instant instance
${i-1}$.

\begin{equation*}
W^{P,1}_{i} = \left( E \cup E_{i} \right) \setminus \left( E \cup E_{i-1} \right) = E_{i} \setminus E_{i-1}
\end{equation*}
\end{axiom}

\begin{axiom}
\label{axiom:continuous datalog: negative window increment}
The window of range 1 at any given instant ${i \in \mathbb{N}}$ over
the stream $N$ entailed by a ${CDP}$ is defined as the set of facts
that were entailed by the state of the ${CDP}$ at the previous instant
instance ${i-1}$ that are no longer entailed by the state of ${CDP}$
at instant $i$.
\begin{equation*}
W^{N,1}_{i} = \left( A' \cup E_{i-1} \right) \setminus \left( A' \cup E_{i} \right) = E_{i-1} \setminus E_{i}
\end{equation*}
\end{axiom}

\begin{axiom}
\label{axiom:continuous datalog: entailment precedes negation}
Facts may only be negated by some part of their justification leaving
the sliding window $W^{S,r}_{i}$, this necessarily being after they
were entailed.  As such, a window of any range ${x \in \mathbb{N}}$
over the stream of entailments $P$ at any given instant ${i \in
  \mathbb{N}}$ is disjoint from a window of any range ${x \in
  \mathbb{N}}$ over the the stream of entailment negations $N$ at any
instant after ${i - X}$

CHECK

\begin{equation*}
\forall X' \geq X \left( W^{P,x}_{i} \cap W^{N,y}_{i-z} = \varnothing \right)
\end{equation*}

\noindent where $z \geq x$.
\end{axiom}

From this, we can see that each instance of an entailment in some
$E_{i}$ may take the same form as an instance of any other fact in a
stream of facts, being $\langle s, p, o \rangle \! : \!t$, where $t$ is
the real time associated with the instant $i_{e}$ in which
$\langle s, p, o \rangle$ becomes newly justified by the ${CDP}$ (more
specifically, by some subset of the instances in $W^{S,r}_{i_{e}}$).
As such, the stream of entailment instances $P$ of a ${CDP}$ takes
the same form as the \emph{streams of facts} already described, thus
producing a concise stream of entailments that may be read by
further stream processing.  In addition, the stream of negations $N$
of a ${CDP}$ also takes this form, but differs from the streams
discussed thus far in that it is not ordered by the entailment time
$i_{e}$ of the entailment instances but by the time at which their
justification \emph{ceases} to be valid (i.e. the first instant
${i_{n} \geq i_{e}}$ in which $W^{S,r}_{i_{n}}$ no longer contains
the subset of fact instances first present in $W^{S,r}_{i_{e}}$ that
justified the entailment instance $\langle s, p, o \rangle \! : \!t$).

Thus far the translation from instantaneous sets of entailments to
streams of entailments has been stated axiomatically as the definition
of those streams.  However, it can be shown from these axioms, as in
\ref{sec:semantics: proof for entailments}, that it is also possible
to reconstruct exactly the instantaneous set $E_{i}$ of entailment
instances of any ${CDP}$ at any given instant from the windows of
sufficient range over its entailed streams $P$ and $N$ at that
instant.

CHECK

\[ \forall Y \geq X \geq R \, \left( E_{i} = W^{P,X}_{i} \setminus W^{N,Y}_{i} \right) \]

The minimum range $x$ is such to achieve the \emph{complete} set of
entailment instances $E_{i}$ of the ${CDP}_{i}$.  In addition, the
minimum range $y$ is such to achieve the \emph{correct} set $E_{i}$.
As such, a \emph{valid window} that contains only and exactly those
instances of entailments in $E_{i}$ at every instant $i$ may be
trivially maintained from the streams $P$ and $N$ continuously
entailed by some ${CDP}$.
\end{nestedsection}
